Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-07-07 paper/ifo-appendix.tex
adding app c, general text editing and tidying

\section{Likelihood of sinusoids in Gaussian noise}
\label{app:sinusoid_likelihood}
\jam{[to do! from Andrew: ``Write a short appendix proving that maximizing the likelihood exp(-n||n/2) with respect to A given a signal A cos (wt + phi) is the Fourier transform. Classic sig proc result useful for teachers to know.'']}

>>>>

\section{Detecting a sinusoidal signal in Gaussian noise}
\label{app:sinusoid_likelihood}
%\jam{[to do! from Andrew: ``Write a short appendix proving that maximizing the likelihood exp(-n||n/2) with respect to A given a signal A cos (wt + phi) is the Fourier transform. Classic sig proc result useful for teachers to know.'']}
In this appendix we demonstrate the the modulus of the Fourier Transform is an appropriate detection statistic when searching for a sinusoidal signal in Gaussian noise. 
We describe the data as
\begin{equation}
x(t) = s(t) + n(t)\,, 
\end{equation}
where $s(t)$ and $n(t)$ are the signal and noise respectively. 
The we describe the signal as 
\begin{equation}
s(t) = A \cos{\omega t + \phi}\,,
\end{equation}
where $A$, $\omega$, and $\phi$ are the amplitude, angular frequency and phase of the signal respectively. 
The log likelihood assuming Gaussian noise is given by
\begin{eqnarray}
\log \mathcal{L} &=& -\frac{1}{2}(n|n)\,, \\
                 &=& -\frac{1}{2}(x-s|x-s)\,, \\ 
                 &=& -\frac{1}{2}(x|x) - \frac{A^2}{2} (\cos[\omega t + \phi] | \cos[\omega t + \phi]) \\ &&+ A (x | \cos[\omega t +\phi])\,, \\ 
                 &=& -\frac{1}{2}(x|x) - \frac{A^2}{4} + A (x | \cos[\omega t +\phi])\,, \label{eqn:logLike}\\ 
\end{eqnarray}
where we define the inner product between $u(t)$ and $v(t)$ as $(u|v) = \int_0^T \mathrm{d} t \, u(t) v(t) $ and $T$ is the total time of the data (and the last line assumes that $T>>1/\omega$ so that $(\cos[\omega t + \phi] | \cos[\omega t + \phi])\approx{1/2}$).

Maximising Eqn.~\ref{eqn:logLike} with respect to $A$ shows that the maximum likelihood is when $A = 2\left( x | \cos[\omega t+\phi]\right)$. 
The maximum likelihood is then, 
\begin{eqnarray}
\log \mathcal{L}_\mathrm{max} = - \frac{1}{2} (x|x) + ( x | cos[\omega t + \phi)^2\,. \label{eqn:maxLogLike}
\end{eqnarray}
Focusing on the second term of Eqn~\ref{eqn:maxLogLike}, we see that the maximum log likelihood of a sinusoidal signal in Gaussian noise is the modulus of the cosine Fourier transform, baring the constant factor $(x|x)$.

### 2020-07-07 paper/ifo-complexAudio.tex
adding app c, general text editing and tidying

% Optical microphone
% aim of optical microphone: play back injected sounds
A natural successor to simple tomes is complex audio, such as music and speech. 
Complex audio signals are not quasi-monochromatic, unlike continuous gravitational wave signals, so the Viterbi algorithm as implemented in Section~\ref{sec:viterbi_wandering} is not applicable directly.
Instead, in this section, we explore how to use the interferometer as an `optical microphone' to detect and play back complex audio signals, using light to capture sound. 
In doing so, we experiment with a hierarchy of passive filters which suppress noise yet make no assumption about the specific form of the signal, unlike the Fourier-based maximum likelihood matched filter tuned to sinusoidal signals in Section~\ref{sec:single_tone} and Appendix~\ref{app:sinusoid_likelihood} \han{to do}. 
The device has precedence in the laser microphones~\cite{laser_microphone} used in the defence industry which operate on a variety of related principles.
It is also analogous to the detection of unmodelled gravitational-wave signals, e.g. broadband bursts from supernovae~\cite{}\han{to do - add citation}. 
%It is also analogous to the detection of unmodelled gravitational wave signals, e.g. broadband bursts from supernovae. [refs; my last sentence can be improved]

@@

The voltage signal from the photo-detector is captured by a MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi Model 3 v1.2~\cite{RaspberryPi:online}, which provides a convenient means to record the photodiode data. 
\han{[from hannah: James does this sound ok? And I have referenced the Raspberry Pi general website, but could you add any appropriate references you used]} \jam{[from James: no problem, sounds good. no obvious references spring to mind that aren't readily available for anyone trying to reproduce the experiment]}
Together, these sample the signal at roughly $16\,{\rm kHz}$.

@@

A Wiener filter is used to compute an estimate of the original signal. 
We label the noisy speech signal sequence as  $\mathbf{x}=\{x(0),\dots, x(N-1)\}$, where $N$ is the length of the data sequence. 
It is the sum of the original clean signal $\mathbf{s}=\{s(0),\dots,s(N-1)\}$, and the noise sequence $\mathbf{w}=\{w(0),\dots,w(N-1)\}$, 
\begin{equation}
    \mathbf{x}=\mathbf{s}+\mathbf{w}
\end{equation}
Given the observation data $\mathbf{x}$, our goal is to estimate $\mathbf{s}$ so that the error between $\mathbf{x}$ and the estimate ${\hat{\mathbf{s}}}$ is minimized in the mean square sense,
\begin{equation}
    \hat{\mathbf{s}}=\mathop{\arg\min}_{\textbf{s}}||\mathbf{x}-\mathbf{s}||_2^2\,. 
\end{equation}
Assuming that $\textbf{x}$ is wide-sense stationary \han{[perhaps a simple definition of wide-sense stationary should be included in brackets here? to do; ask Changrong]} and that $\textbf{w}$ is stationary and uncorrelated with the signal, we represent the $n$th order Wiener filter by its coefficients $\textbf{h}=\{h(0),\dots(n)\}$. 
The famous Wiener-Hopf equation~\citep{noble1959methods} provides us a way to determine $\textbf{h}$, as
\begin{equation}
\label{eqn:winer-hopf}
\begin{bmatrix}  
r_{xx}[0]&r_{xx}[1]&\dots& r_{xx}[n]\\
r_{xx}[1]&r_{xx}[0]&\dots &r_{xx}[n-1]\\
\vdots&\vdots&\ddots&\vdots\\
r_{xx}[n]&r_{xx}[n-1]&\dots &r_{xx}[0]
\end{bmatrix}
\begin{bmatrix}
h[0]\\
h[1]\\
\vdots\\
h[n]
\end{bmatrix}=
\begin{bmatrix}
r_{ss}[0]\\
r_{ss}[1]\\
\vdots\\
r_{ss}[n]
\end{bmatrix}
\end{equation}
where $r_{xx}$ and $r_{ss}$ are the auto-correlation functions of $\mathbf{x}$ and $\mathbf{s}$ between timestep $i$ and $i+n$, 
%\begin{equation*}
%\left\{ 
\begin{eqnarray} 
r_{xx}[n] &~=~& E[x(i)~x(i+n)] \,,\\
r_{ss}[n] &~=~& E[s(i)~s(i+n)] \,.
%r_{xx}[n] &=& E[x(i)\, &x(i+n)] \\ 
%r_{ss}[n] &=& E[s(i)\, &s(i+n)] 
\end{eqnarray} 
%\right\}  
%\end{equation*}
%where $r_{xx}$ and $r_{ss}$ is the auto-correlation function of $x$ and $s$.
If we do not need to guarantee the causality of the Wiener filter, i.e. we can estimate the current signal based on both past and future observations, then Eqn.~\ref{eqn:winer-hopf} can be represented in the frequency domain as
\begin{equation}
\label{eqn:wiener}
    H(f)=\frac{P_{ss}(f)}{P_{xx}(f)}=\frac{P_{ss}(f)}{P_{ss}(f)+P_{ww}(f)}
\end{equation}
where $P_{xx}(f),P_{ss}(f),P_{ww}(f)$ stands for the frequency spectrum of the noisy observation, clean signal, and pure noise respectively. 
Intuitively, from Eqn.~\ref{eqn:wiener}, we can see that the Wiener filter amplifies the input signal where signal to noise ratio (SNR) is high, and weakens the signal where SNR is low.
We note that this filter is non-causal, which may not be suitable for real time signal processing. 
More detailed analysis can be found in Ref.~\citep{10.5555/151045}. 
In this work, we construct a causal Wiener filter based on Eqn.~\ref{eqn:winer-hopf}, and set the length of the filter to $n=100$. 
A larger $n$ provides more smoothing of the input signal, however requires more computational memory. 
The choice of $n=100$ is a reasonably selection to balance smoothing and efficiency.

>>>>

% Optical microphone
% aim of optical microphone: play back injected sounds
A natural successor to simple tomes is complex audio, such as music and speech. 
Complex audio signals are not quasi-monochromatic, unlike continuous gravitational wave signals, so the Viterbi algorithm as implemented in Section~\ref{sec:viterbi_wandering} is not applicable directly.
Instead, in this section, we explore how to use the interferometer as an `optical microphone' to detect and play back complex audio signals, using light to capture sound. 
In doing so, we experiment with a hierarchy of passive filters which suppress noise yet make no assumption about the specific form of the signal, unlike the Fourier-based maximum likelihood matched filter tuned to sinusoidal signals in Section~\ref{sec:single_tone} and Appendix~\ref{app:sinusoid_likelihood}. 
The device has precedence in the laser microphones~\cite{laser_microphone} used in the defence industry which operate on a variety of related principles.
It is also analogous to the detection of unmodelled gravitational-wave signals, e.g. broadband bursts from supernovae~\cite{}\han{to do - add citation}. 
%It is also analogous to the detection of unmodelled gravitational wave signals, e.g. broadband bursts from supernovae. [refs; my last sentence can be improved]

@@

The voltage signal from the photo-detector is captured by a MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi Model 3 v1.2~\cite{RaspberryPi:online}, which provides a convenient means to record the photodiode data. 
\han{[from hannah: James does this sound ok? And I have referenced the Raspberry Pi general website, but could you add any appropriate references you used]} \jam{[from James: no problem, sounds good. no obvious references spring to mind that aren't readily available for anyone trying to reproduce the experiment]} \han{[could you add a couple of useful resources? We should try our best to help out the reader in terms of reproducibility]}
Together, these sample the signal at roughly $16\,{\rm kHz}$.

@@

A Wiener filter is used to compute an estimate of the original signal. 
We label the noisy speech signal sequence as  $\mathbf{x}=\{x(0),\dots, x(N-1)\}$, where $N$ is the length of the data sequence. 
It is the sum of the original clean signal $\mathbf{s}=\{s(0),\dots,s(N-1)\}$, and the noise sequence $\mathbf{w}=\{w(0),\dots,w(N-1)\}$, 
\begin{equation}
    \mathbf{x}=\mathbf{s}+\mathbf{w}
\end{equation}
Given the observation data $\mathbf{x}$, our goal is to estimate $\mathbf{s}$ so that the error between $\mathbf{x}$ and the estimate ${\hat{\mathbf{s}}}$ is minimized in the mean square sense,
\begin{equation}
    \hat{\mathbf{s}}=\mathop{\arg\min}_{\textbf{s}}||\mathbf{x}-\mathbf{s}||_2^2\,. 
\end{equation}
Assuming that $\textbf{x}$ is wide-sense stationary~\footnote{
A random process $\{x(t)\}$ is wide sense stationary (WSS) if for all
$t_1,t_2 \in R$, it satisfies (1) mean is time invariant, i.e.,
$\mu_x(t_1)=\mu_x(t_2)=\text{constant}$; (2) autocorrelation depends
only on time difference, i.e., $R_x(t_1,t_2)=R_x(\tau),\tau=t_1-t_2$.}
and that $\textbf{w}$ is stationary and uncorrelated with the signal, we represent the $n$th order Wiener filter by its coefficients $\textbf{h}=\{h(0),\dots(n)\}$. 
The famous Wiener-Hopf equation~\citep{noble1959methods} provides us a way to determine $\textbf{h}$, as
\begin{equation}
\label{eqn:winer-hopf}
\begin{bmatrix}  
r_{xx}[0]&r_{xx}[1]&\dots& r_{xx}[n]\\
r_{xx}[1]&r_{xx}[0]&\dots &r_{xx}[n-1]\\
\vdots&\vdots&\ddots&\vdots\\
r_{xx}[n]&r_{xx}[n-1]&\dots &r_{xx}[0]
\end{bmatrix}
\begin{bmatrix}
h[0]\\
h[1]\\
\vdots\\
h[n]
\end{bmatrix}=
\begin{bmatrix}
r_{ss}[0]\\
r_{ss}[1]\\
\vdots\\
r_{ss}[n]
\end{bmatrix}
\end{equation}
where $r_{xx}$ and $r_{ss}$ are the auto-correlation functions of $\mathbf{x}$ and $\mathbf{s}$ between timestep $i$ and $i+n$, 
%\begin{equation*}
%\left\{ 
\begin{eqnarray} 
r_{xx}[n] &~=~& E[x(i)~x(i+n)] \,,\\
r_{ss}[n] &~=~& E[s(i)~s(i+n)] \,.
%r_{xx}[n] &=& E[x(i)\, &x(i+n)] \\ 
%r_{ss}[n] &=& E[s(i)\, &s(i+n)] 
\end{eqnarray} 
%\right\}  
%\end{equation*}
%where $r_{xx}$ and $r_{ss}$ is the auto-correlation function of $x$ and $s$.
If we do not need to guarantee the causality of the Wiener filter, i.e. we can estimate the current signal based on both past and future observations, then Eqn.~\ref{eqn:winer-hopf} can be represented in the frequency domain as
\begin{equation}
\label{eqn:wiener}
    H(f)=\frac{P_{ss}(f)}{P_{xx}(f)}=\frac{P_{ss}(f)}{P_{ss}(f)+P_{ww}(f)}
\end{equation}
where $P_{xx}(f),P_{ss}(f),P_{ww}(f)$ stands for the frequency spectrum of the noisy observation, clean signal, and pure noise respectively. 
Intuitively, from Eqn.~\ref{eqn:wiener}, we can see that the Wiener filter amplifies the input signal where signal to noise ratio (SNR) is high, and weakens the signal where SNR is low.
We note that this filter is non-causal, which may not be suitable for real time signal processing. 
More detailed analysis can be found in Ref.~\citep{10.5555/151045}. 
In this work, we construct a causal Wiener filter based on Eqn.~\ref{eqn:winer-hopf}, and set the length of the filter to $n=100$. 
A larger $n$ provides more smoothing of the input signal, however requires more computational memory. 
The choice of $n=100$ is a reasonably selection to balance smoothing and efficiency.

### 2020-07-07 paper/ifo-singleTone.tex
changes to transfer function flowchart
adding app c, general text editing and tidying

%The resulting data is a video capture of the whole interference pattern. 
%As only the frequency of the interference ring motion is needed, w
We take data from a single off-centre pixel indicated by the red dot in Fig.~\ref{fig:interference_pattern}. 
The green-channel of the video is used as an approximation to the total intensity as the laser produces monochromatic green light (at $532\,{\rm nm}$).
As an initial inspection, we play the timeseries of the intensity using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
The two tones sound similar by ear and don't produce any noticeable beating. 
The result is quantified in Fig.~\ref{fig:webcam_spectrum}. 
A tone of $2.09\,{\rm Hz}$ is played through the speaker and the data recorded via webcam. 
Figure~\ref{fig:webcam_spectrum} shows the Fourier transform of the recording.
We measure a peak amplitude at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$. 
Two harmonics can also be seen in Fig.~\ref{fig:webcam_spectra} at integer harmonics of the main peak. 
The $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ peaks have amplitudes of $0.103$ and $0.098$ respectively as a fraction of the height of the main peak. 
They are probably due to the nonlinear system response discussed in Section~\ref{} and Appendix~\ref{}.

>>>>

%The resulting data is a video capture of the whole interference pattern. 
%As only the frequency of the interference ring motion is needed, w
We take data from a single off-centre pixel indicated by the red dot in Fig.~\ref{fig:interference_pattern}. 
The green-channel of the video is used as an approximation to the total intensity as the laser produces monochromatic green light (at $532\,{\rm nm}$).
As an initial inspection, we play the timeseries of the intensity using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
The two tones sound similar by ear and don't produce any noticeable beating. 
The result is quantified in Fig.~\ref{fig:webcam_spectrum}. 
A tone of $2.09\,{\rm Hz}$ is played through the speaker and the data recorded via webcam. 
Figure~\ref{fig:webcam_spectrum} shows the Fourier transform of the recording.
We measure a peak amplitude at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$. 
Two harmonics can also be seen in Fig.~\ref{fig:webcam_spectrum} at integer harmonics of the main peak. 
The $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ peaks have amplitudes of $0.103$ and $0.098$ respectively as a fraction of the height of the main peak. 
They are probably due to the nonlinear system response discussed in Section~\ref{sec:ifo} and Appendix~\ref{app:intensity_derivation}.

### 2020-07-07 paper/ifo-wanderingTone.tex
adding app c, general text editing and tidying

One method to search for a wandering signal is to split the data into several shorter segments which are analysed individually. 
The detection statistics in every segment are combined to form a grid in time and frequency (spectrogram). 
The value at each spectrogram point determines the likelihood that a signal is present at a particular time and frequency. 
In gravitational-wave data analysis, a detection statistic is used to calculate the likelihood of a signal given the antenna beam pattern of the detector, which varies with as the Earth rotates and orbits the Sun~\cite{JKS:1998}.
The detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}). 
In this work the detection statistic is the amplitude (modulus) of the Fourier transform of the data at each time segment, which maximizes the likelihood of detecting a sinusoid in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood} \han{to do} \jam{[makes sense but is beyond me]}.
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin. 
The task is then to find the most likely wandering frequency signal through the spectrogram. 
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
It is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

>>>>

One method to search for a wandering signal is to split the data into several shorter segments which are analysed individually. 
The detection statistics in every segment are combined to form a grid in time and frequency (spectrogram). 
The value at each spectrogram point determines the likelihood that a signal is present at a particular time and frequency. 
In gravitational-wave data analysis, a detection statistic is used to calculate the likelihood of a signal given the antenna beam pattern of the detector, which varies with as the Earth rotates and orbits the Sun~\cite{JKS:1998}.
The detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}). 
In this work the detection statistic is the amplitude (modulus) of the Fourier transform of the data at each time segment, which maximizes the likelihood of detecting a sinusoid in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. 
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin. 
The task is then to find the most likely wandering frequency signal through the spectrogram. 
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
It is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.
